Governing through patronage: the bargain for education in decentralized Brazil

### 2022-08-29 paper/chapter_education.tex
fix: incorporate latest version of chapter

\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1.5in, bottom = 1.5in, left = 1.5in, right = 1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{hyperref}

@@

\DeclareMathOperator*{\argmax}{arg\,max}
\onehalfspacing
\usepackage[hang,flushmargin]{footmisc}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}
\newcommand{\bfv}{\mathbf{v}}
\newcommand{\blue}{\textcolor{blue}}
% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}

\bibliographystyle{unsrtnat}

\title{Governing through patronage: the bargain for education in decentralized Brazil}
\date{\today}

\author{Galileu Kim}
\affil{Princeton University}

\begin{document}

\maketitle

\begin{abstract}
    How can governments improve the quality of education? I argue that local politicians govern through patronage, undermining education at the expense of voters. Building on qualitative evidence in Brazil, I leverage large-scale administrative data to show that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering them positions into the educational sector. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. Weak electoral backlash suggest that patronage is primarily a political elite game, with limited accountability to the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

\newpage
\section{Introduction}

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains woefully low: functional illiteracy remains widespread, basic mathematical operations.\footnote{The World Bank calls this ``learning poverty". See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}. Clearly, politicians play an important role in delivering these services \citep{gulzar_politicians_2017,besley_political_2003}, but what are the incentives they face and what are their implications for bureaucratic institutions responsible for delivering them. \citep{cochrane_policies_1983, falleti_decentralization_2010}. How can local governments improve the quality the of public services they deliver? In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning?

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, children cannot perform arithmetic operations. How can public education be improved?

Previous studies have found that certain incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on individualized interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But how do these actors relate to one another in local institutions? Who holds power over public education, and whose interests is it meant to serve?

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

In this paper, I argue that public education in municipalities across Brazil have been captured by political elites, with personnel appointments into public schools and management dominated by bargains between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over voter welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The estimation proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of specifications and different measures of turnover.

@@

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, but I highlight the non-electoral policy goals of executive leaders. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.

The paper is structured as follows. Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. Section \ref{sec:context} presents the context and data. In section \ref{sec:theory}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model. Section \ref{sec:empirics} presents the research design and main results. Section 5 concludes.

@@

In this section I review extant literature on public goods provision and the politics underlying it, focusing on more recent studies of bureaucratic personnel and political leaders reshape these institutions. I also highlight how my research incorporates multiplicity in political actors and how this affects bargaining over public sector jobs. I address this gap by adapting previous analyses of executive-legislative bargaining to bureaucratic control at the local level.

@@

Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. Toral 2019 finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. \citet{gulzar_politicians_2017} show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. \citet{akhtari_political_2015}, on the other hand, highligh the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

@@

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle_jobs_2012, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy. \citet{raile_executive_2011, power_optimism_2010}.

In the Brazilian federal context, executive-legislative relations are analyzed under the prism of presidential coalitionism. Executive leaders garner legislative support from the National Congress by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, neto_presidential_2006, figueiredo_executivo_1999}.

In municipalities, mayors have to garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, "city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

@@

Bureaucracies exposed to turnover experience productivity shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citet{gailmard_slackers_2007}. Focusing on education, studies show that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher \citet{clotfelter_teacher_2007}. \citet{akhtari_political_2015} finds that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citet{colonnelli_patronage_2017}.

Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers. In the following section I describe the institutional context for public education in Brazil, as well as the data employed for the the estimation.

\section{Institutional context and data}
\label{sec:context}

@@

In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}. In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and substraction. The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school strudents attend a public school. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/descriptive_enrollment.pdf}
    \caption{\textbf{Student enrollment by administration.}}
    \label{fig:enroll}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/descriptive_staff_breakdown}
\end{figure}

@@

@@

Brazil collects fine-grained data on education and makes it publicly available for research. For this study, data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute fosr Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.\linewidth]{plots/descriptive_staff_turnover.pdf}
    \label{fig:staff_turnover}
    \caption{Proportion of public sector jobs by department. Note that I only keep the top 5 categorie}
\end{figure}

I propose three different specifications for measuring bureaucratic turnover. From the RAIS I obtain the percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

$$\text{turnover}_{sjt} = \frac{\text{exit}_{sjt} + \text{entry}_{sjt}}{N_{sjt} + N_{sjt-1}}$$

\blue{Insert more information and descriptive paragraph on turnover index here}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\linewidth]{plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the elecoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

@@

\section{Theory}
\label{sec:theory}

@@

The government $G$ and opposition $O$ compete over legislative votes to enact their preferred policies.\footnote{Note that for this model, I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition. In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators. The total amount of political resources available is $W_G$ and $W_O$, which for the mayor includes public sector appointments.

@@

We solve the game through backward induction. The timing of the game is as follows:

@@

Following Groseclose and Snyder (1996) and Banks (2000) we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$. Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed. The mayor then solves

@@

I impose the following two assumptions:
\begin{align*}
    A1 &: v_{(n+1)/2} < 0\\
    A2 &: v_1 < 2W_B/(n+1) 
\end{align*}
$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$. Therefore $A$ must bribe at least one voter. $A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win. $A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.

>>>>

\documentclass[12pt]{report}
\usepackage{setspace}
\usepackage[top = 1in, bottom = 1in, left = 1.5in, right = 1.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\usepackage{multirow}
\usepackage{titling}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{kbordermatrix}
\usepackage{kotex}
\usepackage{epigraph}
\usepackage{sectsty}
\allsectionsfont{\singlespacing}

@@

\DeclareMathOperator*{\argmax}{arg\,max}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{definition}{Definition}

\newcommand{\wb}{\overline{w}}
\newcommand{\pb}{\overline{p}}
\newcommand{\bb}{\overline{b}}
\newcommand{\de}{\delta}
\newcommand{\St}{\mathcal{S}}
\newcommand{\Sb}{\overline{\mathcal{S}}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\note}{\textcolor{blue}}
\renewcommand{\theproposition}{\arabic{proposition}}
\newcommand{\real}{\mathbb{R}_+^n}
\newcommand{\bfv}{\mathbf{v}}
\newcommand{\blue}{\textcolor{blue}}
% \renewcommand{\includegraphics}{\includegraphics[width=0.6\linewidth]}

\bibliographystyle{apsr}

\section*{Abstract:}

\emph{How do political institutions shape decisions over bureaucratic personnel? And what are the implications of these management decisions on public service delivery? I argue that local mayors govern through patronage, lowering the quality of educational services. Qualitative evidence gathered in fieldwork in Brazil, as well as large-scale administrative data suggest that public education is captured by local political elites. Mayors buy off local legislators to enact policy agendas by offering positions in the department of education. The degree to which patronage occurs varies: when mayors have a stronger ally base in the city council, they face less pressure to bargain. Patronage induces turnover in educational staff, with negative downstream effects on student learning. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.}

\section{Introduction}

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy remains widespread, basic mathematical operations inaccessible.\footnote{The World Bank calls this ``learning poverty". See \url{https://www.worldbank.org/en/topic/education/brief/learning-poverty}}. Politicians play an important role in deciding whether or not to deliver these services \citep{gulzar_politicians_2017,besley_political_2003}, but alongside elected officials, bureaucratic institutions shape how these services are delivered \citep{pepinsky_bureaucracy_2017, finan2017personnel}. How do politicians manage bureaucratic institutions, and what are their downstream consequences for public services?

Previous studies have found that individual-level incentives, electoral or performance related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on targeted interventions, these studies provide compelling evidence that political and bureaucratic actors respond to policy changes.\citep{finan_personnel_2015} But across the developing world, political and bureaucratic actors do not govern alone. Public officials are embedded in institutions that shape and constrain their actions \citep{thelen_historical_1999}. Moreover, it is common that instead of a single political actor, multiple principals compete over the same pool of resources, requiring coalitional bargains to effectively govern \citep{francois_how_2015}. How do these coalitional bargains affect patronage appointments, and what are its implications for the population they are meant to serve?

To answer these questions, I conducted fieldwork research in municipalities across Brazil over the course of two years. In interviews with local politicians, school principals and teachers, I uncovered the pervasive influence of clientelism, and how it extends to public education with negative consequences on educational outcomes \citep{stokes2013brokers}. Clientelism in the educational sector was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government that occur between elections \citep{raile_executive_2011}. In Brazilian municipalities, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

In this paper, I argue that public education in municipalities across Brazil is a bargaining chip between political elites, with personnel appointments into public schools and management dominated by transactions occurring between executive and legislative branches. Similar to coalition building in presidential contexts, mayors bargain for legislative support for their policy agenda through the allocation of public sector positions to city councilors and their constituencies \citep{laver_coalitions_1990, power_optimism_2010}. This effectively crowds out electoral accountability, as mayors prioritize building support by other political elites over constituency welfare \citep{ferejohn_incumbent_1986}. By catering to the city council through patronage appointments, mayors shuffle the local educational bureaucracy, with negative downstream consequences for public school students.

The empirical identification of these theoretical mechanisms proceeds in two parts. First, I show that political alignment between mayors and city councilors has a direct effect on patronage. Building on a canonical model of legislative vote-buying, I propose a theoretical model to analyze patronage under an institutionalized separation of powers, deriving my main empirical test: mayors who face less opposition in the city council engage in less patronage. To verify this hypothesis, I build a set of indicators to track educational staff turnover, leveraging administrative data of over 2 million school principals and teachers employed by municipalities. In line with qualitative accounts and theoretical expectations, mayors who have a stronger ally base in the legislature engage in less patronage. This results holds across a set of model specifications and different measures of turnover.

@@

This study contributes to an emerging literature on the politics of personnel and public services \citep{pepinsky_bureaucracy_2017, finan_personnel_2015}. I analyze how incentives shape political decision on how to manage local bureaucracies \citep{duflo_incentives_2012, gulzar_politicians_2017}, highlighting the policy goals of executive leaders and how these require bargains that leave public service beneficiaries worse off. I also contribute to a growing research agenda on the consequences of political competition, demonstrating that increased political fragmentation between executive and legislative branches can lead patronage and decrease in the quality of public services. \citep{gottlieb_countervailing_2019, ferraz_corrupting_2012}. I bring to focus the end-to-end provision of public services, echoing a long-standing literature on state capacity \citep{kohli_state-directed_2004, evans_embedded_1995}.

The paper is structured as follows. Section \ref{sec:literature} provides an overview of the scholarly debate over public goods provision and personnel, as well as more specific treatments of the subject in Brazil. Section \ref{sec:context_edu} presents the context and data. In section \ref{sec:theory_bargain}, I present the main argument, with a formal treatment of the subject using a canonical vote-buying model. Section \ref{sec:empirics_edu} presents the research design and main results. Section \ref{sec:conclusion_edu} concludes.

@@

@@

Recent studies have added nuance to these claims, finding that high bureaucratic performance can coexist with political interference. \citet{toral2019benefits} finds in Brazil that school principals appointed by mayors tend to perform better than their non-appointed counterparts in standardized test scores. \citet{gulzar_politicians_2017} show that local politicians who are able to internalize electoral benefits make bureaucrats exert more effort, increasing local employment. \citet{akhtari_political_2015}, on the other hand, highlight the pitfalls of political capture, showing that party turnover can lead to the replacement of school principals, with detrimental effects for student learning.

@@

For every elected mayor in Brazil, a group of legislators are also elected into office. These political actors have competing claims over the local bureaucracy, with important implications for public service delivery. This structure parallels other settings with an institutinalized separation of powers and a bureaucratic pie to be split among the actors \citep{grindle2012jobs, mccarty_appointments_2004}. Divergent political interests can lead to strategic interaction between executive and legislative actors. A rich literature in Brazil explores these relations, with important insights to how executive and legislators bargain over bureaucracy \citep{raile_executive_2011, power_optimism_2010}.

In the Brazilian federal context, executive-legislative relations are predominantly analyzed under the prism of presidential coalitionism \citep{neto_presidential_2006}. Due to a multiparty system, that is common across federal, state and municipal level, rarely do presidents, governors or mayors secure a majority in the legislative chamber. Executive leaders garner legislative support from the legislature by exchanging key positions in the federal bureaucracy, appointing members of their legislative coalition into cabinet positions. \citep{raile_executive_2011}. In a setting characterized by weak party cohesion and programmatic commitments \citet{ames_electoral_1995, lucas_ideological_2010}, bureaucratic positions for members of the coalition provide material incentives for legislators to support the executive agenda \citet{batista_o_2013, figueiredo_executivo_1999}.

In municipalities, mayors have to, in a similar fashion, garner legislative support from city councilors to secure budgetary approval and implement desired policies. Due to weak programmatic commitments at the local level, public sector jobs are used to legislative support.\footnote{Interview with C, August and September 2019.} Mayors enjoy full discretion into how to appoint workers into the public sector, and use patronage to coopt legislative support from members of the coalition, a practice known locally as \textit{empreguismo}. As noted by a former mayor of the municipality of Sobral, ``city councilors knocked on my door with a list of names for people they wanted me to hire."\footnote{Interview with C, August 2019.} These hires induce changes in personnel, with important consequences for bureaucracies and educational services.

@@

Bureaucracies exposed to turnover experience performance shocks, often with detrimental effects. As new staff enters the bureaucracy, they must learn and acquire skills to deliver services to the population \citep{gailmard_slackers_2007}. \citet{clotfelter_teacher_2007} finds that students taught by inexperienced teachers perform worse than those attending class with an experienced teacher. \citet{akhtari_political_2015} show, in municipalities in Brazil, that students attending a school with a recently appointed school principal perform worse in standardized test scores. When bureaucratic turnover is driven by patronage, political concerns take precedence over meritocratic ones \citep{colonnelli_patronage_2017}.

This empirical record suggests that, as in any organization, bureaucratic turnover is often costly, especially in the initial phases of change. While the benefits of changes have been extolled by the embrace of New Public Management in more recent years \citep{osborne_reinventing_1993}, an attention to continuity and stability in bureaucratic organizations has been the subject of constant attention in scholarly literature on public administration \citep{o2003plus,suleiman_dismantling_2013}. Bureaucratic organizations are complex, and have to fulfill mandates that require familiarity with both the population to be served and the procedures set in place to comply both with external and internal regulations.

If disruption is to be beneficial, there must be time for these turnover induced changes to mature and bear fruits. This extended time horizon has to be coupled with an identification of organizational priorities, rather than political ones \citep{nunes_gramatica_1997}. If cycles of disruption are associated with political imperatives rather than bureaucratic performance ones, an additional misalignment of incentives takes hold of bureaucratic institutions. \citet{evans_embedded_1995} posits that high bureaucratic performance is premised on embedded autonomy. What that theoretical and empirical claim implies is that while bureaucracies must respond to changes in their environment -- and alter themselves in the process -- they also require a degree of autonomy to effect these gradual changes.

\section{Institutional context and data}
\label{sec:context_edu}

@@

In Brazil, the responsibility to provide public, primary education is primarily delegated to local governments \citet{paschoal_historia_2009}. In this critical learning period, students acquire skills such as reading/writing, as well as foundational concepts in math such as addition and subtraction. The municipal educational system has increased in relevance over the past few decades, and currently over 60 percent of lower school students attend a public school. Figure \ref{fig:enroll} plots the total number of students in primary education per government level, including private schools. As of 2016, over 25 million students were enrolled in over 115 thousand municipal schools.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_enrollment.pdf}
    \caption{\textbf{Student enrollment by administration.}}
    \label{fig:enroll}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown}
\end{figure}

@@

\subsection{Bureaucratic turnover}

In interviews with local education officials, it became clear the extent to which turnover was detrimental to the provision of public services. In conversations with the \emph{n\'{u}cleo gestor} (administrative board) of a school in the municipality of I, administrators complained of the need to learn the tasks and processes required to run a school. This process, according to them, took over a year or even two, and if the administrative board or teachers were replaced, the administrative expertise of the school was reset and the new administration had to go through the entire process once again.

In the municipality of I, one of the poorest run educational systems in the state of Cear\'{a}, bureaucratic turnover was the rule, not the exception. As put

Negotiations between executive and the legislative thus have a knock-on effect on the quality of educational services, as political considerations lead to bureaucratic turnover at the school and administrative level. In this study, the primary focus is on bureaucrats working within the boundaries of a school: school principals and teachers.

@@

Brazil collects fine-grained data on education and makes it publicly available for research. Data on education are collected from two main sources: the SAEB and School Census.\footnote{These can be accessed \hyperlink{http://portal.inep.gov.br/web/guest/dados}{here}.} The SAEB (National System for Educational Assessment) is a biannual standardized exam administered by the INEP (National Institute for Educational Studies and Research) to all public schools and a sample of private schools. In 2017, over 5 million students, in 5th and 9th grade, took the exam, testing their proficiency in both Mathematics and Portuguese. In this study, test scores are the primary metric for assessing the quality of education received by students.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/saeb_map.pdf}
    \caption{Uneven quality of municipal education. Polygons represent municipalities, warmer colors indicate higher municipal average in standardized test scores. Averages are weighted by student participation rate.}
    \label{fig:saeb_map}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/descriptive_staff_breakdown.pdf}
    \caption{Proportion of public sector jobs by department}
    \label{fig:staff_turnover}
\end{figure}

I propose three different specifications for measuring bureaucratic turnover. From the RAIS I calculate the municipal-level percentage of teachers and principals who are dismissed and hired at any given year. Using school census data, it is possible to track individual teachers across time and schools. With that data I calculate a turnover index for school $s$, grade $g$, in municipality $j$ at year $t$, based on an index proposed by \citet{pereira_junior_indicadores_2016}. The number of teachers leaving and entering the school at a given year are summed and divided by the total number of teachers in the current and previous period.

$$\text{turnover}_{sgjt} = \frac{\text{exit}_{sgjt} + \text{entry}_{sgjt}}{N_{sgjt} + N_{sgjt-1}}$$

The following example illustrates how the index operates. Suppose that in school $s$ for the year 2010, there are three teachers working. In the following year, one new teacher arrives and two leave. Our index would reflect these changes: $\text{turnover} = \frac{1 + 2}{3 + 2} = 0.6$. If, in contrast, no teachers left or entered the school, we would instead have a turnover index of 0 for the school. The added advantage of this measure is that it simultaneously captures a broader definition of bureaucratic turnover, that includes both exits and entries into the local school. Figure \ref{fig:hist_turnover} presents the distribution of turnover index over time.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_turnover_index.pdf}
    \caption{Distribution of turnover index by year, pooled across schools. Note that for robustness, only schools with more than 5 teachers are included.}
    \label{fig:hist_turnover}
\end{figure}

Note that there is not significant variation over time of the turnover index, indicating that there are no visible time trends or reduction in teacher turnover. This indicates that turnover is a chronic challenge for schools in Brazil and has not reduced significantly over time, despite efforts by the Ministry of Education to increase professionalization of eduactional staff through more strict public exam requirements. It is important to highlight that entry or exit in this context does not indicate that a teacher is necessarily leaving or entering the educational system, but rather being relocated to another school within the same system.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.6\linewidth]{chapters/chapter_1/plots/hist_coalition.pdf}
    \caption{Distribution of share of legislative seats allied with the mayor during the electoral campaign for the eletcoral period of 2002-2016.}
    \label{fig:hist_coalition}
\end{figure}

@@

\section{Theory}
\label{sec:theory_bargain}

In this section, I present an abridged version of the theoretical framework presented by \citet{banks_2000_buying} and \citet{groseclose_1996_buying}. In this framework, two players -- government and opposition -- compete over legislative votes in order to pass their preferred policy proposals. Each legislative member casts an aye or nay vote depending on both their a priori preferences over the competing proposals as well as the bribe schedule induced by the competition between the two players. There is perfect information: both parties can fully observe the actions of each other and there is no uncertainty in the model.

This simplification costs us a degree of realism -- one could imagine that bribes are actually occurring in the backstage -- but gains us parsimony and tractability. The key comparative static of interest is how exogenous changes in the preference profile of legislative members induce changes in the bribing decisions of the government. Intuitively, if a legislative member opposes a proposal, they require additional bribes to vote in favor. Generalizing, as the preference profile of members of the legislature are more favorable (opposed) towards the policy proposal of the politician, the less (more) resources must the politician spend to garner their vote.

@@

In this setting, two players, the government $G$ and opposition $O$, compete over legislative votes to enact their preferred policies.\footnote{Note that for this model, I use the terms \emph{mayor} and \emph{government} interchangeably.} There are two possible outcomes: a policy $x$ favored by the government, and the status quo, denote as $y$, preferred by the opposition. In order to implement her policy the mayor must pass a simple majority vote in the city council, comprised of an odd number $N$ of legislators. The total amount of political resources available is $W_G$ and $W_O$, which in our context are simply public sector appointments for the mayor, and other kinds of resources -- monetary, limited appointments -- for the opposition.

@@

The game is solved through backward induction. The timing of the game is as follows:

@@

Following \citet{groseclose_1996_buying} and \citet{banks_2000_buying} we focus our analysis on the set of equilibria in which the mayor wins.\footnote{Since strategies for both players are symmetrical, any set of equilibria in which the mayor loses can be modeled as cases in which the the opposition loses.} In this context, the amount of patronage resources $W_G$ is sufficiently large relative to $W_O$ and $\bfv$ that the mayor's preferred policy $x$ is implemented over $y$. Let $U(\bfv, W_O)$ denote the set of unbeatable patronage schedules for the mayor, and for any patronage schedule let $S(a) = \sum_{i = 1}^n a_i$ be the total amount of patronage disbursed. The mayor then solves

@@

Following \citet{banks_2000_buying}, I impose the following two assumptions:
\begin{align*}
    A1 &: v_{(n+1)/2} < 0\\
    A2 &: v_1 < 2W_B/(n+1) 
\end{align*}
$A1$ implies that absent any bribes by $A$, $y$ will defeat $x$. Therefore $A$ must bribe at least one voter. $A2$ implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win. $A2$ also implies that for all $a \in U^l(\mathbf{v}, W_B)$, it must be that $t(a) \geq 2W_B/(n + 1)$, otherwise $B$ can bribe a majority from $C(a)$ itself and win the vote.

These assumptions allow us to restrict our analysis to unbeatable monotonic leveling schedules, which we denote as $U_m^l$.\footnote{For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$. Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$. Then, under $A2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$: $a'_i = t(a) - v_i$, $a'_j = 0$ and $a'_m = a_m$ for all $m \notin i, j$. Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e. $a'_i \leq a_j$. Also, because of $A2$ $a'_j \leq a_j$, since it guarantees $a_j$ is non-negative. Generalizing, and recalling that $v_1 \geq ... \geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$. Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters. Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.} We can simplify the total expenditure on patronage by the government, $S(a)$, as
